
# Extract data within a boundary {#sanctuary}
>notebook filename | 02-sanctuary.Rmd    
history | converted to R notebook from sanctuary.R  

In this exercise, you will download data from within the boundaries of the Monterey Bay National Marine Sanctuary (MBNMS) and visualize the data in a map. 

The exercise demonstrates the following skills:  

* Loading data from a CSV file  
* Using **rerddap** to retrieve information about a dataset from ERDDAP 
* Using the **rxtractogon** function to extract satellite data within an polygon over time  
* Mapping satellite data  

## Install packages and load libraries
```{r install,message=FALSE,warning=FALSE}

# Function to check if pkgs are installed, install missing pkgs, and load
pkgTest <- function(x)
{
  if (!require(x,character.only = TRUE))
  {
    install.packages(x,dep=TRUE)
    if(!require(x,character.only = TRUE)) stop(x, " :Package not found")
  }
}

# create list of required packages
list.of.packages <- c("ncdf4", "parsedate", "rerddap", "sp", "devtools", "ggplot2", "RColorBrewer", 
                      "reshape2", "maps", "mapdata", "jsonlite")

# Run install and load function
for (pk in list.of.packages) {
  pkgTest(pk)
}

# create list of installed packages
pkges = installed.packages()[,"Package"]

# Check if devtools pkgs are install. Install missing pkgs.
if(!('rerddapXtracto' %in% pkges)) {
  devtools::install_github("rmendels/rerddapXtracto")}
if(!('plotdap' %in% pkges)) {
  devtools::install_github('ropensci/plotdap')} 
if(!('rerddap' %in% pkges)) {
  devtools::install_github("ropensci/rerddap")}

library(rerddap)
library(rerddapXtracto)
library(mapdata)
```

## Load sanctuary boundary coordinates
For this example, the sanctuary boundary is definded within a comma separated file (mbnms.txt) in the shapes folder. The file contains a series of longitude and latitude coordinates the together draw the the boundry of the sanctuary on a map, like tracing a dot-to-dot drawing.  

>Longitude, Latitude  
-120, 52.0  
-120, 52.0  
-120, 52.0  
'...  
-120, 52.0  

A boundary file for Olympic Coast National Marine Sanctuary (OCNMS) is also availaible in the shapes folder. Additional sanctuary boundaries may be obtained at   [http://sanctuaries.noaa.gov/library/imast_gis.html](http://sanctuaries.noaa.gov/library/imast_gis.html).

**The script below:**   

* Loads the boudary data from the CSV file into a dataframe  
* Extracts the longitude and latitude data into vector variables  


```{r load_csv}
# Assign the csv path and file to the boundary variable
boundary <- 'shapes/mbnms.txt'
# boundary <- 'shapes/ocnms.txt'  # uncomment this line and comment out the above line to run for the OCNMS 

# Load CSV boudary data into the dataframe "poly"
poly <- read.csv(boundary)
names(poly) <- c("Longitude","Latitude")

# Extract the longitude and latitude data into vector variables
xcoord <- poly$Longitude
ycoord <- poly$Latitude
```
## Select the chloropyll dataset  
For this example we will use 750 m VIIRS monthly chlorophyll dataset (ID erdVHNchlamday)  

**The script below:**  
* Gathers information about the dataset (metadata) using **rerddap**  
* Displays the information  
```{r dataInfo}
# Use rerddap to get dataset metadata 
dataInfo <- rerddap::info('erdVHNchlamday')  # N. Pacific 750 m VIIRS chl
# Display the metadata 
dataInfo
```
## Set the options for rxtractogon  
* Use the name of the chlorophyll parameter that was displayed above in dataInfo: **parameter <- "chla"** 
* The metadata from dataInfo also shows you that this variable has a altitude coordinate that equals zero. Set the value of the time coordinate to zero: **zcoord <- 0.**  
* The time variable passed to xtracogon must contain two elements, the start and endpoints of the desired timeperiod. 
  * This example makes use of the "last" option of ERDDAP to retrieve data from the most recent time step.  
  * The "last' option also accepts the "-" operator. To request the time step with the second morst recent data use "last-1". 
  * In the script below the time variable (tcoord) is defined as **tcoord <- c("last-1", "last")**  

```{r options}
# set the parameter to extract
parameter <- 'chla'
# set the time range
tcoord <- c("last-1", "last")
# Assign longitude and latitude vectors from the CSV file to variables
xcoord <- mbnms$Longitude
ycoord <- mbnms$Latitude
# set the altitude variable to zero
zcoord <- 0. 
```
## Extract data and mask it using rxtractogon  
* Set the arguments within the **rxtractogon** function using the dataset information (dataInfo), the longitude (xcoord) and latitude (ycoord) vectors from the CSV file, the time (tcoord) vector and alitude (zcoord) variable. 
* Then run **rxtractogon** to extract data from the "erdVHNchlamday" dataset and mask out data not falling within the MBNMS boundary.  
* List the data
```{r octogon}
## Request the data
sanctchl <- rxtractogon (dataInfo, parameter=parameter, xcoord=xcoord, ycoord=ycoord,tcoord=tcoord,zcoord=zcoord)

## List the returned data
str(sanctchl)
```

The extracted data contains two time periods of chlorophyll data within the sanctuary boundaries. In this example we will make a plot of the second time period. 

##Prepare data for plotting
```{r plot}
## Remove extraneous zcoord (altitude) dimension from chlorophyll data
sanctchl$chla <- drop(sanctchl$chla) 

## Define a helper function mapFrame to reshape the data to be used in ggplot
mapFrame<- function(longitude,latitude,chla){
  dims<-dim(chla)
  chla<-array(chla,dims[1]*dims[2])
  longitude<-longitude-360
  chlaFrame<-expand.grid(x=longitude,y=latitude)
  chlaFrame$chla<-chla
  return(chlaFrame)
}
```
### Plot the data
* Create a map that shows the satellite data within the sanctuary bounds
* Latitude and longitude are the axes
* Chl data is sanctchl$chla variable
* Uncomment the png and dev.off lines to save the plot to a file

```{r map}
#set the map extent
xlim = c(min(xcoord)+.001*min(xcoord),max(xcoord)-.001*max(xcoord))
ylim = c(min(ycoord)-.01*min(ycoord),max(ycoord)+.01*max(ycoord))

# png(file = "Sanctuary.png")  # uncomment to save the plot
mbnmsFrame<-mapFrame(sanctchl$longitude+360,sanctchl$latitude,log(sanctchl$chla[,,2]))
w <- map_data("worldHires", ylim = ylim, xlim = xlim)
myplot<-ggplot() + geom_path(data=poly,aes(x=Longitude,y=Latitude),colour="black")   
myplot<-myplot + 
  geom_raster(data = mbnmsFrame, aes(x = x, y = y, fill = chla),interpolate = FALSE) + 
  theme_bw() + 
  scale_fill_gradientn(colours =rev(rainbow(12)),limits=c(-1.5,3)) +
  ylab("Latitude") + xlab("Longitude") +
  geom_polygon(data = w, aes(x=long, y = lat, group = group), fill = "grey80") +
  coord_fixed(1.3,xlim = xlim, ylim = ylim) + 
  ggtitle(paste("log(Chla)",sanctchl$time[2])) 
myplot

# dev.off()  # uncomment to save the plot
```

```{r, nice-fig2, echo=FALSE}
htmltools::img(src = knitr::image_uri("images/cw_logo_notext.png"), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px; width:100px')
```
