# Extract data within a boundary {#sanctuary}
>notebook filename | 02-sanctuary.Rmd  
history | converted to R notebook from sanctuary.R  

In this exercise, you will download data from within the boundaries of the Monterey Bay National Marine Sanctuary (MBNMS) and visualize the data in a map. 

The exercise demonstrates the following skills:  

* Using **rerddap** to retrieve information about a dataset from ERDDAP 
* Using the **rxtractogon** function to extract satellite data within an polygon over time  
* Mapping satellite data  

## Install packages and load libraries
```{r install,message=FALSE,warning=FALSE}
pkges = installed.packages()[,"Package"]
# Function to check if pkgs are installed, install missing pkgs, and load
pkgTest <- function(x)
{
  if (!require(x,character.only = TRUE))
  {
    install.packages(x,dep=TRUE)
    if(!require(x,character.only = TRUE)) stop(x, " :Package not found")
  }
}

# create list of required packages
list.of.packages <- c("ncdf4","parsedate","rerddap","plotdap","sp","rerddapXtracto",
                      "ggplot2", "RColorBrewer", "colorRamps","cmocean", 
                      "reshape2", "maps", "mapdata", "jsonlite")

# Run install and load function
for (pk in list.of.packages) {
  pkgTest(pk)
}

# create list of installed packages
#pkges = installed.packages()[,"Package"]

```

## Load sanctuary boundary coordinates
The **rerddapXtracto**  package comes with the dataset **mbnms** which conatains the longitude and latitude values for the boundary of the Monterey Bay National Marine Sanctuary.  These coordinates draw the the boundary of the sanctuary on a map, like tracing a dot-to-dot drawing.  Take a quick look at the contents of this data variable. 
 
```{r loadmbnms}
str(mbnms)
```

Additional sanctuary boundaries may be obtained at   [http://sanctuaries.noaa.gov/library/imast_gis.html](http://sanctuaries.noaa.gov/library/imast_gis.html).

**The script below:**   

* Extracts the longitude and latitude data into vector variables  

```{r latlon}

# Extract the longitude and latitude data into vector variables
xcoord <- mbnms$Longitude
ycoord <- mbnms$Latitude
```
## Select the chloropyll dataset  
For this example we will use 750 m VIIRS monthly chlorophyll dataset (ID erdVHNchlamday)  

**The script below:**  
* Gathers information about the dataset (metadata) using **rerddap**  
* Displays the information  
```{r dataInfo}
# Use rerddap to get dataset metadata 
dataInfo <- rerddap::info('erdVHNchlamday')  # N. Pacific 750 m VIIRS chl
# Display the metadata 
dataInfo
```
## Set the options for rxtractogon  
* Use the name of the chlorophyll parameter that was displayed above in dataInfo: **parameter <- "chla"** 
* The metadata from dataInfo also shows you that this variable has a altitude coordinate that equals zero. Set the value of the time coordinate to zero: **zcoord <- 0.**  
* The time variable passed to xtracogon must contain two elements, the start and endpoints of the desired time period. 
  * This example makes use of the "last" option of ERDDAP to retrieve data from the most recent time step.  
  * The "last' option also accepts the "-" operator. To request the time step with the second most recent data use "last-1". 
  * In the script below the time variable (tcoord) is defined as **tcoord <- c("last-1", "last")**  

```{r options}
# set the parameter to extract
parameter <- 'chla'
# set the time range
tcoord <- c("last-1", "last")
# Assign longitude and latitude vectors from the CSV file to variables
xcoord <- mbnms$Longitude
ycoord <- mbnms$Latitude
# set the altitude variable to zero
zcoord <- 0. 
```
## Extract data and mask it using rxtractogon  
* Set the arguments within the **rxtractogon** function using the dataset information (dataInfo), the longitude (xcoord) and latitude (ycoord) vectors from the CSV file, the time (tcoord) vector and altitude (zcoord) variable. 
* Then run **rxtractogon** to extract data from the "erdVHNchlamday" dataset and mask out data not falling within the MBNMS boundary.  
* List the data
```{r octogon}
## Request the data
sanctchl <- rxtractogon (dataInfo, parameter=parameter, xcoord=xcoord, ycoord=ycoord,tcoord=tcoord,zcoord=zcoord)

## List the returned data
str(sanctchl)
```


## Select the time period    

The extracted data contains two time periods of chlorophyll data within the sanctuary boundaries. In this example, the second time period will be selected. 

```{r chl}

sanctchl1 <- sanctchl
sanctchl1$chla <- sanctchl1$chla[, , 2]
sanctchl1$time <- sanctchl1$time[2]

```
## Plot the data

Below we will demonstrate two different ways of plotting the data: 

(1) plotBBox - Using the **plotBBox** function in rerddapXtracto allows you to quickly plot the data

(2) ggplot- Using ggplot allows you to customize the map, you should use either base graphics or ggplot.  Here we use ggplot to make a map of the data. 

### Using the plotBBox function   

```{r map}
plotBBox(sanctchl1,plotColor='algae',maxpixels=100000)

```

**Apply a math function to the data**  

Often it better to plot the log of Chlorophyll data, so here we will apply the logrithmic function to the chlorophyll data 

```{r log}
myFunc <- function(x) log10(x) 
plotBBox(sanctchl1, plotColor = 'algae',maxpixels=100000, myFunc=myFunc)


```

** Using ggplot   

First reformat the data for use in ggplot by flatten the XY data array into a long-format vector of length X*Y.  
Next, expand the longitude and latitude vectors into vectors of length X*Y   


```{r flatten}
dims <- dim(sanctchl1$chla)
sanctchl.lf <- expand.grid(x=sanctchl1$longitude,y=sanctchl1$latitude)
sanctchl.lf$chla <- array(sanctchl1$chla,dims[1]*dims[2])
```

**Determine the limits for the bounding box**  
```{r limits}
xlim <- c(min(xcoord),max(xcoord))
ylim <- c(min(ycoord),max(ycoord))
```


**Get the coastline data for the region**  

Note: for a large area it is better to use **world** rather than **worldHires** base map.
```{r mapbase}
coast <- map_data("worldHires", ylim = ylim, xlim = xlim)
```


**Make the map**   
```{r map1}
ggplot(data = sanctchl.lf, aes(x = x, y = y, fill = chla)) +
  geom_tile(na.rm=T) +
  geom_polygon(data = coast, aes(x=long, y = lat, group = group), fill = "grey80") +
  theme_bw(base_size = 15) + ylab("Latitude") + xlab("Longitude") +
  coord_fixed(1.3,xlim=xlim,ylim=ylim,expand = c(0, 0)) +
  ggtitle(unique(as.Date(sanctchl1$time))) +
  scale_fill_gradientn(colours = c(blue2red(12),"Black"), oob = scales::squish,
                       limits=c(0,25), na.value="white") 
```






```{r, nice-fig2, echo=FALSE}
htmltools::img(src = knitr::image_uri("images/cw_logo_notext.png"), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px; width:100px')
```
