# Matchups to ship or animal tracks {#xyt-matchup}
>notebook filename | 03-xyt_matchup.Rmd    
history | converted to R notebook from xyt_matchup.R  

This exercise you will map a track, like that produced by the following the position of an animal, ship, or glider, using longitude, latitude, and time coordinates. 

The exercise demonstrates the following techniques:  

* Using the **rxtracto** function to extract satellite data along a track  
* Using **rerddap** to retrieve information about a dataset from ERDDAP  
* Using **plotTrack** to plot the satellite data onto a map as well as to make an animation 
* Loading data from a tab separated file  
* Plotting the satellite data onto a map

This data is taken from the ERDDAP server at [http://coastwatch.pfeg.noaa.gov/erddap/](http://coastwatch.pfeg.noaa.gov/erddap/)

## Install required packages and load libraries
```{r install,message=FALSE,warning=FALSE}

# Function to check if pkgs are installed, and install any missing pkgs

pkgTest <- function(x)
{
  if (!require(x,character.only = TRUE))
  {
    install.packages(x,dep=TRUE)
    if(!require(x,character.only = TRUE)) stop(x, " :Package not found")
  }
}

# create list of required packages
list.of.packages <- c("ncdf4", "parsedate","rerddap","plotdap","sp","rerddapXtracto",
                      "graphics", "maps", "mapdata","RColorBrewer","ggplot2",
                      "cmocean","gifski","png")

# create list of installed packages
pkges = installed.packages()[,"Package"]

# Install and load all required pkgs
for (pk in list.of.packages) {
  pkgTest(pk)
}

```
## Get XYZ coordinates
In this exercise we will the "Marlintag38606" dataset, which is included in the "rerddapXtracto" package. The "Marlintag38606" dataset contains the XYT (longitude, latitude, time) coordinates of a tagged marlin in the Pacific Ocean (courtesy of Dr. Mike Musyl of the Pelagic Research Group LLC).   

The "Marlintag38606" file has this structure:  

```{r}
str(Marlintag38606)
```

We will use the "lon", "lat" and "date" variables to get the matching satellite data. Here the time variable is already in a date format.  Often when reading in your own data you will have to convert the date into a date format (if formating the date remember R syntax is Y for a 4 digit year and y for a 2 digit year)

```{r}

# For convenience make shorter names for the variables  
xcoord <- Marlintag38606$lon  
ycoord <- Marlintag38606$lat
tcoord <- Marlintag38606$date

```

## Select the dataset and download its metadata  
For this example use the SeaWiFS 8-day composite chlorophyll dataset (ID erdSW2018chla8day)  

**The script below:**  

* Gathers information about the dataset (metadata) using **rerddap**  
* Displays the information  

**Set the following arguments for rerddap**  

* Set the dataset ID: dataset <- 'erdSW2018chla8day'

* The default source ERDDAP for **rerddap** is "https://upwell.pfeg.noaa.gov/erddap". Since we are pulling the data from the ERDDAP at "http://coastwatch.pfeg.noaa.gov/erddap/", change the url to url = "http://coastwatch.pfeg.noaa.gov/erddap/"

```{r dataInfo}
dataset <- 'erdSW2018chla8day'
# Use rerddap to get dataset metadata 
# if you encouter an error reading the nc file clear the rerrdap cache: 
rerddap::cache_delete_all(force = TRUE)
dataInfo <- rerddap::info(dataset, url= "https://coastwatch.pfeg.noaa.gov/erddap/")
# Display the metadata
dataInfo
```
## Extract the satellite data  

* Double check dataInfo to make sure the dataset covers the time, longitude, and latitude ranges in your XYT data.  

* Use the name of the chlorophyll parameter that was displayed above in dataInfo: **parameter <- "chlorophyll"**  

* Use the xcoord, ycoord, and tcoord vectors you extracted from the marlin tag file. 

* Some datasets have an altitude dimension. If so, then zcoord must be included in the rxtracto call. The "erdSW2018chla8day" dataset does not include an altitude dimension.  

* Define the search "radius" for the gridded data. The **rxtracto** function allow you to set the size of the box used to collect data around the track points using the xlen and ylen arguments. The values for xlen and ylen are in degrees. For our example use 0.2 degrees for both arguments. Note: You can also submit vectors for xlen and ylen, as long as the arethe same length as xcoord, ycoord, and tcoord

* Run the rxtracto function to extract the data from ERDDAP.  

```{r rxtracto}

parameter <- 'chlorophyll'

xlen <- 0.2 
ylen <- 0.2

# Some datasets have an altitude dimension. If so, then zcood must be included in the rxtracto call.  
# If the dataInfo shows an altitude dimension, uncomment "zcoord <- 0" and include tcoord=tcoord in the rxtracto call.
# zcoord <- 0.

swchl <- rxtracto(dataInfo, 
                  parameter=parameter, 
                  xcoord=xcoord, ycoord=ycoord, 
                  tcoord=tcoord, xlen=xlen, ylen=ylen)

```

After the extraction is complete, "swchl" will contain the following columns.

```{r}
str(swchl)
```


## Plotting the results

Two different ways of plotting the data are demonstrate below: 
(1) using the "plotTrack" function in rerddapXtracto, which is designed specifically to plot the results from "rxtracto".   
(2) using ggplot, which allows customization of the plots  


### Using plotTrack 
This example uses a color palette specifically designed for chlorophyll. 
  
```{r plot}
# Uncomment the png line and the dev.off() line to save the image
# png(file="xyt_matchup.png")

plotTrack(swchl, xcoord, ycoord, tcoord, plotColor = 'algae')

# Uncomment the png line and the dev.off() line to save the image
#dev.off()
```

**Animating the track** 

To make a cumulative animation of the track:

```{r}
plotTrack(swchl, xcoord, ycoord, tcoord, plotColor = 'algae',
                    animate = TRUE, cumulative = TRUE)
```

### Using ggplot.  

Put the data into a dataframe for use with ggplot.  Subtract 360 from the longitude values to get them ontp values expressed from -180 to 180. 

```{r}
swchl.df <- data.frame(x=xcoord-360,y=ycoord, t=tcoord, chl=swchl$`mean chlorophyll`)
```


**Set up the limits for the bounding box**  

The bounding box limits should be larger then the minimum and maximum values of xcoord and ycoord  

```{r}
xlim <- c(-160, -100)
ylim <- c(18,27)
```



**Get the coastline data for the region**

If its a big area, use world, rather than worldHires
```{r mapbase1}
coast <- map_data("worldHires", ylim = ylim, xlim = xlim)

```

**Map the data**

```{r mapdata, warning=FALSE}
ggplot(swchl.df,aes(x,y)) + 
  geom_point(aes(color=chl)) + 
  geom_polygon(data = coast, aes(x=long, y = lat, group = group), fill = "grey80") +
  theme_bw(base_size = 15) + ylab("Latitude") + xlab("Longitude") +
  coord_fixed(1.3,xlim=xlim,ylim=ylim,expand = c(0, 0)) +
  scale_color_gradientn(colours = brewer.pal(n = 9, name = "Greens"), oob=scales::squish,
                        limits=c(0,.2), na.value="gray") 
```


**Try this on your own**
This match up was done using weekly (8-day) data. Try rerunning the example using the daily (erdSW2018chla1day) or the monthly (erdSW2018chlamday) satellite data product and see how the results differ. 

## Crossing the Dateline
In July 2019 version 0.4.1 of "reddapXtracto"" was updated allowing "rxtracto"" to work on data that crosses the dateline. In this example we will extract chlorophyll data for a grid of stations along the Aleutian Islands. 

__Create an station array__   
For crossing the dateline the longitudes for that animal/ship track must be in 0-360 format.  
* Create a grid of stations from 172E to 170W (190°) and 50-54N, spced every 2°.
* Then, set up vectors with these values, and then make arrays of the station longitudes and latitudes

```{r station}
lat <- seq(50,54,2)
lon <- seq(172,190,2)

stax <- matrix(lon,nrow=length(lat),ncol=length(lon),byrow=TRUE)
stay <- matrix(lat,nrow=length(lat),ncol=length(lon),byrow=FALSE)
```

To input values into "rxtracto" the longitudes and latitudes need to be in vector format
```{r inputvar}
xcoord <- as.vector(stax) 
ycoord <- as.vector(stay) 
```

Define the search "radius" in the x any y directions, in units of degrees
```{r radius}
xlen <- 0.2 
ylen <- 0.2 
```

Create an array of dates. For this exercise we are going to assume all stations were sampled in the same month, so we are going to make all the values the same, but they don't have to be.   

```{r array1}
tcoord <- rep('2019-04-15',length(xcoord))
```

Selects the dataset and parameter for the extraction  
In this example the dataset chosen is the monthly NOAA VIIRS chlorophyll data 
```{r select1}
dataset <- 'nesdisVHNSQchlaMonthly'
dataInfo <- rerddap::info(dataset)
parameter <- 'chlor_a'
```

**Look at DataInfo to see if dataset has an altitude dimension.** 

```{r alt}
dataInfo
```

Since this dataset has an altitude dimension, supply an altitude parameter in the "rxtracto" call 

```{r addalt}
zcoord <- 0.*xcoord
```

Make the call to match up satellite data with station locations.  
```{r matchup}
chl <- rxtracto(dataInfo, 
                  parameter=parameter, 
                  xcoord=xcoord, ycoord=ycoord, zcoord=zcoord,
                  tcoord=tcoord, xlen=xlen, ylen=ylen)
```


Next map out the data. Will do this two different ways, using base graphics and using "ggplot".  "plotTrack", the routine used in the example above, is part of the "rerddapXtracto" package, and is designed to easily plot the output from "rxtracto", but currently it can not handle crossing the dateline, so it cannot use it for this example. 

### Make a map using base graphics  

First set up the color palette. 
This will use a yellow-green palette ("speed") from the cmocean package  


```{r base}
cols <- cmocean('speed')(9)
chlcol <- cols[as.numeric(cut(chl$'mean chlor_a',breaks = 9))]
```

**Identify stations which have a satellite values**  

```{r idst}
gooddata <- !is.na(chl$'mean chlor_a')
```

**Set-up the layout to have a map and a color bar**   

```{r setlayout}
oldmar <- par("mar")
layout(t(1:2),widths=c(6,1))
par(mar=c(4,4,1,.5))
```

Create the base map, and then overlay stations with data, and then overlay empty circles around all statons 
```{r base2}
ww2 <- map('world', wrap=c(0,360), plot=FALSE, fill=TRUE)
map(ww2, xlim = c(140, 240),ylim=c(45,70), fill=TRUE,col="gray80",lforce="e")
map.axes(las=1)

points(xcoord[gooddata],ycoord[gooddata],col=chlcol, pch=19, cex=.9)
points(xcoord,ycoord, pch=1, cex=.9)
```

**Add the colorbar** 
```{r cbar}
par(mar=c(4,.5,5,3))
chlv <- min(chl$'mean chlor_a'[gooddata])+(0:9)*(max(chl$'mean chlor_a'[gooddata])-min(chl$'mean chlor_a'[gooddata]))/10
image(y=chlv,z=t(1:9), col=cols, axes=FALSE, main="Chl", cex.main=.8)
axis(4,mgp=c(0,.5,0),las=1)
```


### Make a map using ggplot graphics  

ggplot handles colorbars much easier than base graphics! 

**Put station lat, long and chl values into a dataframe for passing to ggplot** 
 
```{r df1}
chlsta <- data.frame(x=xcoord,y=ycoord,chl=chl$'mean chlor_a')
```

**Get land boundary data in 0-360 units of longitude** 

```{r land}
mapWorld <- map_data("world", wrap=c(0,360))
```

**Make the map** 
```{r map8}
ggplot(chlsta) +
  geom_point(aes(x,y,color=chl)) +
  geom_polygon(data = mapWorld, aes(x=long, y = lat, group = group)) + 
  coord_cartesian(xlim = c(140,240),ylim = c(47,70)) +
  scale_color_gradientn(colours=cmocean('speed')(8)) +
  labs(x="", y="")
```



      


```{r, nice-fig2, echo=FALSE}
htmltools::img(src = knitr::image_uri("images/cw_logo_notext.png"), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px; width:100px')
```