# TurtleWatch {#turtlewatch}
>notebook filename | 06_turtlewatch_xtracto.Rmd  
history | converted to R notebook from R_Turtlewatch_rXtracto.R  

## Install required packages and load libraries 

```{r install, message=FALSE, warning=FALSE}

# Function to check if pkgs are installed, install missing pkgs, and load
pkgTest <- function(x)
{
  if (!require(x,character.only = TRUE))
  {
    install.packages(x,dep=TRUE)
    if(!require(x,character.only = TRUE)) stop(x, " :Package not found")
  }
}

list.of.packages <- c( "ncdf4", "RCurl",  "raster", "colorRamps", "maps", "mapdata",
                       "ggplot2", "RColorBrewer")

# create list of installed packages
pkges = installed.packages()[,"Package"]

for (pk in list.of.packages) {
  pkgTest(pk)
}

# Check if devtools pkgs are install. Install missing pkgs.
if(!('rerddapXtracto' %in% pkges)) {
  devtools::install_github("rmendels/rerddapXtracto")}
if(!('plotdap' %in% pkges)) {
  devtools::install_github('ropensci/plotdap')} 
if(!('rerddap' %in% pkges)) {
  devtools::install_github("ropensci/rerddap")}

library(rerddap)
library(plotdap)
library(rerddapXtracto)
```

## Select the Satellite Data
* Use the MUR SST dataset (ID jplMURSST41mday)  
* Gather information about the dataset (metadata) using **rerddap**  
* Displays the information  
```{r dataInfo}
# CHOOSE DATASET and get information about it 

dataInfo <- rerddap::info('jplMURSST41mday')
parameter <- 'sst'
```

## Get Satellite Data  
* Select an area off the coast of California: longitude range of -130 to -115 east  and latitude range of 25 to 40 north  
* Set the time range to a single day: tcoord=c('2018-06-06','2018-06-06'))  

```{r getdata}

# latitude and longitude of the vertices
ylim<-c(25,40)
xlim<-c(-130,-115)

# Choose an area off the coast of California
# Extract the data
SST <- rxtracto_3D(dataInfo,xcoord=xlim,ycoord=ylim,parameter=parameter, 
                   tcoord=c('2018-06-06','2018-06-06'))

# Drop command needed to reduce SST from a 3D variable to a 2D  one  
SST$sst <- drop(SST$sst) 
```
## Make a quick plot
```{r qplot}
image(SST$sst, col=matlab.like2(255))  
```
## DEFINING THE THERMAL NICHE OF LOGGERHEAD TURTLES
* TurtleWatch define the optimal temmal range of Loggerhead turtle to be 17.5-18.5 degrees C.  

```{r temp}
## Define turtle temperature range
min.temp <- 17.5
max.temp <- 18.5
```
## Prepare data for plotting
```{r plotprep}
## create another variable where habitat temperaure set to NA 
SST$sst2 <- SST$sst
SST$sst2[SST$sst2 >= min.temp & SST$sst2 <= max.temp] <- NA

# Restructure the data frame to polot using ggplot, create a function to do this 

mapFrame<- function(longitude,latitude,sst){
dims<-dim(sst)
sst<-array(sst,dims[1]*dims[2])
sstFrame<-expand.grid(x=longitude,y=latitude)
sstFrame$sst<-sst
return(sstFrame)
}

# Restructure the data 

sstFrame<-mapFrame(SST$longitude,SST$latitude,SST$sst2)

```
## Plot the Data
```{r plot}
coast <- map_data("worldHires", ylim = ylim, xlim = xlim)
my.col <- colorRampPalette(rev(brewer.pal(11, "RdYlBu")))(22-13) 

par(mar=c(3,3,.5,.5), las=1, font.axis=10)

myplot<-ggplot(data = sstFrame, aes(x = x, y = y, fill = sst)) +
  geom_raster(interpolate = FALSE,na.rm=T) +
  geom_polygon(data = coast, aes(x=long, y = lat, group = group), fill = "grey80") +
  theme_bw(base_size = 15) + ylab("Latitude") + xlab("Longitude") +
  coord_fixed(1.3,xlim = xlim, ylim = ylim) +
  ggtitle(unique(as.Date(SST$time))) +
  scale_fill_gradientn(colours = rev(rainbow(12)),limits=c(10,25),na.value = "firebrick4") 

myplot

```

```{r, nice-fig2, echo=FALSE}
htmltools::img(src = knitr::image_uri("images/cw_logo_notext.png"), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:10px; width:100px')
```